The Folly of the Crowd: \\ Market Inefficiencies in Peer to Peer Lending

### 2015-03-30 draft.Rnw
Clarified Adj.Roi comments. Fixed VK suggestions.

\documentclass[12 pt]{article}
\usepackage{graphicx, verbatim}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage[linktoc=all]{hyperref}
\usepackage{titlesec}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}
\begin{document}
\setcounter{tocdepth}{2}
\linespread{1.3}
\tableofcontents

@@

The early auction regime is closest to Hayek's vision of the free market, since
interest rates are determined freely by investors through auctions, with few
limitations on which borrowers they can invest in.
I find that it is precisely in this free-market regime that the three
predictions of the EMH are contradicted.
In the two later regimes, when Prosper first provides much more guidance to investors, and then actually sets interest rates, investor returns improve dramatically.
The evidence across regimes is broadly inconsistent with the view that local knowledge and the wisdom of the crowd improves price discovery.
To the contrary, it appears that a "centralized' Prosper mechanism of either guiding or setting interest rates actually leads to fewer predictable mispricings: as Prosper becomes more involved in the rate-setting process, the excess returns available using publicly available information approach 0, while aggregate returns exhibit higher mean and lower variance.

@@

Under all three regimes, lenders could choose to whom to lend and how much to
lend, allowing them to diversify across many loans, thereby decreasing the
variance of their returns. Before a potential borrower's listing is posted on
the website, Prosper authenticates their social security number, driver's
license and address, and pulls the borrower's credit history from Experian,
which includes credit score, total number of delinquencies, current
delinquencies, and many other historical credit variables.\footcite{freedman} 
In exchange, In exchange for verifying borrowers' identities and managing the
marketplace, Prosper charges a 0.5-3 percentage
point origination fee on each loan depending on the borrower's risk profile and
loan duration. 
This fee,usually 1\%, is added to the borrower's monthly payment.
Lenders are also charged a servicing fee of 0.5-1\%. There are no prepayment penalties.
This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.\newline

In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.\newline
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of Loans funded on Prosper in each month of the sample. The second panel shows the total dollars lent to borrowers in each month of the sample.', fig.height=4>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup), Dollars = sum(LoanOriginalAmount))
g2 = melt(g2,id.vars = c("mgroup",'regime'))

@@

<<catsTab, cache=FALSE>>=
catTab = filter(nd, purpose != "NA") %>% count(purpose, sort=TRUE) %>% mutate(Frequency = n/sum(n))
xtable(catTab, digits=3, label='catTab', caption = "Stated Reasons for Borrowing on Prosper (NAs removed)")
@

@@

Second, whereas previously Prosper used only credit score to determine a borrower's Credit Grade (eg. AA), the site began offering a more advanced, multi variable formula to determine each borrower's credit grade for investors consideration. Finally, the site also partnered with FolioFN, a secondary market where lenders could liquidate their positions. Crucially, Prosper still had no direct role in setting interest rates, investors still used an auction, but with the extra information credit grades that more closely matched default probabilities. 
Besides credit score, Prosper began using (and still uses) the following ``Key variables" to determine its revamped credit grades: \footnote{CITE =  Estimated Loss Page}
\begin{itemize}
\item  TotalInquiries: The number of times potential lenders requested the borrowers' credit portfolio from the credit bureau, which is usually Experian or a competitor.
\item CurrentDelinquencies: The number of delinquent loans associated with the borrower.
\item Credit card utilization: how much of available credit the borrower uses on a monthly basis.
\item Number of open credit card accounts
\item Debt to income ratio
\item Loan payment performance on prior Prosper loans.
\end{itemize}
This Auction rate-setting mechanism continued for more than a year after the shutdown, until December 20, 2010.
\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 21, 2010, lenders no longer determine the loan rate in an auction. Instead, Prosper uses the intelligent credit grades given to investors in the late auction regime to set interest rates on loans, with slightly less stringent under-writing standards to determine eligibility. The change was motivated by a desire to accelerate the listing process, so that investors could reinvest money more quickly. Moreover, towards the end of the late auction regime, auctions were arriving at interest rates that were almost completely determined by the credit score. Under the fixed-rate regime, investors choose whether or not to invest in a given loan, at the interest rate set by Prosper, so bids only consist of a dollar amount.\newline

\section{Literature}
\subsection{Efficient Markets Hypothesis}
In layman's terms, the efficient markets hypothesis (EMH), states that 'it is impossible to beat the market.' More specifically, Fama explains that market prices should always fully reflect publicly available information. Therefore, EMH rules out the possibility of strategies that earn excess returns using only publicly available information. \footcite{fama}
In the P2P lending context, this market price is the interest rate determined by investors (or accepted by investors in the fixed-rate regime), and paid by the borrower, and returns are the cash flows received from borrowers divided by the amount invested.
(Fama, 1970) notes the dependence of most tests of market efficiency on a model of risk and return.
The models, including most famously, Sharpe's capital asset pricing model ('CAPM'), make two predictions:
First,  since an investor can always hold cash, the average return on risky investments must be positive. Second, securities with higher systematic risk earn higher average returns.\footcite{sharpe}

@@

Freedman and Jin (2008) are more focused on the failures of the auction regime,
Although the main focus of their paper is the role of soft information and
social networks on Prosper, they also study how investors learn to weight these
different variables, and confirm the finding of negative expected returns in the
early auction regime: ``lenders, especially those that joined
Prosper early, did not fully understand the market risk. We estimate, on
average, Lenders would have expected an internal rate of return of -.62\% to
1.38\% if they had correctly understood the risk distribution of Prosper
loans.''``Vigorous'' lender learning, they argue, explains the increase in interest
rates and borrower creditworthiness, and associated increase in lender
returns.\footcite{freedman}

@@

Rather differently than Hayek, I find that when investors are left on their own
to set prices, markets are not nearly as efficient as the prices set by ``a
single mind'', that aggregates the data. Furthermore, as interest rate setting
becomes more centralized, interest rates are more sensitive to the hard credit variables
that actually predict default.
t and other unused citations can be found in hypo.cut.tex

@@

\subsection{Variable Definitions for the Prosper Data}
<<sumTable,cache=FALSE>>=
nd$default.rate =  nd$defaulted
k4 = c('roi','default.rate', 'LenderYield','BorrowerRate','regime','LoanOriginalAmount','Investors','CreditScoreEst','DebtToIncomeRatio', 'InquiriesLast6Months')
ag4 <- aggregate(.~regime, nd[k4], function(x) c(mean = mean(x), sd = sd(x)))
d4 = t(do.call(data.frame, c(ag4, check.names = FALSE)))
d5 = data.frame(d4)
names(d5) = d4[1,]
d6 = d5[-1,]
xtable(d6,caption='Means and Standard Deviations for Relevant Variables by Regime.',digits=3, label='sumStats')
@
For each loan, Prosper offers 63 interesting variables, I now define an important subset.
\begin{itemize}
\item BorrowerRate: Annual interest rate paid by the borrower
\item TotalCreditLinespast7years: Number of credit lines, including cards and loans, used by the borrower in the seven years before the prosper loan was originated.
\item Inquirieslast6Months: Number of times a lender inquired about the Borrower's credit report, in the 6 months before the loan was originated. Traditional lenders make inquiries when a borrower applies for a loan, so this serves as a proxy for the number of times someone has borrowed in the past 6 months.
\item CurrentDelinquencies: Number of loans on which the borrower is delinquent at the origination date.
\item DelinquenciesLast7Years: The number of loans which the borrower has not repaid fully over the 7 years before the origination date.
\item MonthlyLoanPayment: The borrower's scheduled monthly payment on the Prosper loan.
\item CreditScoreRangeLower: The lower bound of the credit score range (which
  bounds his raw credit score) , that bounds a borrower's credit grade, as
  provided by Experian. Prosper provides the range instead of the raw score to
  preserve borrower anonymity.
\item CreditScoreEst: The average of the lower and upper bounds of the credit score range. 
\item PublicRecordsLast10Years: Number of public records of borrowing activity in the past 10 years at the origination date. 
\item DebtToIncomeRatio: The debt to income ratio of the borrower at the origination date. This value is capped at 10.01 (any debt to income ratio larger than 1000\% will be coded as 1001\%).
\item TotalProsperPaymentsBilled: Number of on time payments the borrower made on Prosper loans at the time they created the listing. This value is 0 if the borrower had no prior loans.
\item LenderYield: The lender's Yield on the loan. Equal to the Interest rate minus servicing fee.
\item BorrowerRate: The borrower's interest rate for the loan.
\item Investors: The number of investors that funded the loan.
\item *.sc: The Variable * was scaled to have 0 mean and unit variance. (found only in regression results)
\end{itemize}

@@

I do not discount or annualize ROIs, because in the period of low interest
rates, the difference between default and repayment is large enough that a few
percentage points don't matter: loans that default average -46\% ROI, loans that
repay average + 24\% ROI. Furthermore, my ROI results are very similar as Freedman
and Jin's, who use discounting.\footcite{freedman}

\subsection{Prosper's Charged-Off vs. Default Terminology}
\footnote{remove this part?}
After the shutdown, Prosper began using the term ``Charged-Off" to mean that a loan is delinquent past 120 days. We observe more defaults than charge-offs before the shutdown, but more charge-offs than defaults after, likely because of the shift in terminology. According to Prosper's website:
\begin{quote}
In general, a debt or account is considered “charged off” when it is unlikely that further payments will be received. Debts are usually charged off after they remain unpaid for a period of time (e.g., 90 to 180 days). Prosper uses the 120 days as the charge off threshold because loans that become over 120 days past due are eligible for sale to a debt buyer, and we have found that there is a steep drop-off in likelihood of further payments after 120 days of delinquency.
\footnote{\url{http://blog.prosper.com/2008/09/16/charge-offs-explained/}}
\end{quote}

@@

I find the following: (1) the mean ROI of early auction regime loans is negative, (2) in the early auction regime, average ROI is lower for higher risk loans, (3) interest rates are insufficiently sensitive to risk in the early auction period and (4) investors made sub-optimal choices, given the information they had at their disposal, and (5) Prosper's interventions improved investor returns.

These findings suggests that although investors were extraordinarily unlucky in
the early auction regime, they made avoidably bad decisions that hurt their
returns. More specifically,
Finding (1) is inconsistent with the EMH prediction that risky assets like Prosper loans should offer investors a positive expected return.\footnote{Otherwise, investors would just hold cash.}
Finding (2) is inconsistent with the prediction that riskier loans should have higher expected returns to compensate investors for taking risks.
Finding (3) describes the sub-optimal investor decisions that led to (1) and (2), in response to arguments that investors chose an unlucky time period.
Finding (4) is inconsistent with Hayek's notion that a free market will arrive at accurate prices without the meddling of a central planner.

@@

ggplot(rateDiff, aes(x=CreditScoreEst, y=meanRoi,size=num,color=regime))+
  geom_line() +
  theme_bw() +
  xlim(500,900) + 
  scale_y_continuous(labels=percent) +
  labs(title='Estimated Credit Score vs. ROI',x= 'Estimated Credit Score'
       , y='Mean ROI',color='Regime',size='Number of Loans')
#Total Inquiries, other cluster vars.
@

Figure \ref{fig:roivscore} shows that higher risk loans earn lower returns in the early auction regime, inconsistent with the EMH prediction of a risk, reward tradeoff.
The expected pattern is shown in later two regimes' curves: riskier loans earn higher average returns. Figure \ref{fig:roivscore} also reveals early auction regime returns are worse than the other two regimes' returns for each bucket, but that the gap is much greater for riskier loans. Again, time period may be a confounding factor, as riskier investments often perform worse in bad economic times, a possibility I discuss later.
We now examine default rates and interest rates, the two features that explain the variation inreturns.

\subsection{Interest Rates and Defaults Across regimes}
<<ratevtime,cache=FALSE,fig.height=3, fig.cap="The blurred lines show the median interest rate of loans originated each month. The solid lins represent the median rate for loans to borrowers in the 640-660 credit score bucket for each month. Loans funded after the shutdown are associated with higher rates after controlling for borrower credit score bucket.">>=
ggplot() +
  geom_line(aes(x=mmeans$mgroup, y=mmeans$mRate, size=mmeans$numLoans, color=mmeans$regime),alpha=.5) + 
  geom_line(aes(x=mm3$mgroup, y=mm3$mRate, size=mm3$numLoans, color=mm3$regime)) + 
  theme_bw() +
  labs(title='Interest Rates Over Time (read caption)', color='Regime',size='Number of Loans',x='Origination Month',y='Mean Rate')+
  scale_y_continuous(label=percent, limits=c(.1, .33))
@

@@

<<defscore.tab,fig.height=3,cache=FALSE,fig.pos='H'>>=
a1 = filter(d2, regime=='Auction-Early')[c('CreditScoreEst','dr')]
colnames(a1) = c('CreditScoreEst','Auction-Early')
a2 = filter(d2, regime=='Auction-Late')[c('CreditScoreEst','dr')]
colnames(a2) = c('CreditScoreEst','Auction-Late')
a3 = filter(d2, regime=='Prosper-rate')[c('CreditScoreEst','dr')]
colnames(a3) = c('CreditScoreEst','Prosper-rate')
ds.tab = left_join(a1, inner_join(a2,a3, by='CreditScoreEst'), by='CreditScoreEst') %>% filter(CreditScoreEst > 600 & CreditScoreEst < 880)
colnames(ds.tab)[1] = 'Credit Score Bucket'
xtable(ds.tab, label='ds.tab', caption='Default rates for borrowers in each credit score for each regime. Credit score buckets are denoted by their midpoint. EG, the 600-619 bucket is in the first row, 609.5. Each column then represents the default rates borrowers in the (column name) regime, in the (row-name) credit bucket. Default rates are much higher for the early auction regime, especially among risky borrowers. This data underlies the previous figure. Early auction regime borrowers in buckets below 500 were removed to save space, but have even higher default rates.')
@
Lower sensitivity to credit scores, and total disregard for differences in
credit score below 600 did not work out well for early auction regime investors,
the relationship between credit score and default probability was even stronger
in that regime than in the other regimes. 
Figure \ref{fig:defvscore} presents the relationship between credit score bucket
and default rate, and shows that the probability of default decreases
(increases) with higher (lower) credit score, for all three regimes.
In the early auction regime, a 100 point increase (5 buckets) in credit score is
associated with a 15\% reduction in default probability. In the late auction
regime, a 100 point increase is associated with a 7\% reduction in default
probability. In the Prosper-rate regime, an 100 point increase is associated
with a 12\% increase in default probability. The data that underly Figure \ref{fig:defvscore}
can be viewed in tabular form in Table \ref{ds.tab}.

@@

Adversaries may argue that since we do not know Propser's exact grading formula, it may include information that was not previously available to borrowers.Table \ref{ratefit2} addresses this argument, by running the same procedure as Table \ref{ratefit}, but with only credit score bucket as an independent variable. The direction of the results is the same: Prosper's change to more informative grading did not only lead to a larger role for those grades in investors' interest rates, but increased the weight of credit score bucket, a variable that investors had access to in the same format in both periods. If the crowd of investors were as wise as Hayek theorizes, Prosper's repackaging of multiple variables into a more informative credit grade should not affect the weight of hard information in investors' risk estimates. Furthermore, Figure \ref{fig:defvscore} makes is clear that investors would have experienced fewer defaults, and thus earned higher returns, if they had used credit buckets more heavily in their risk assessments, given the poor returns associated with low credit borrowers. We can thus interpret some of the poor returns of the early auction regime to a lower level of reliance on hard information.

>>>>

\documentclass[12 pt]{article}
\usepackage{graphicx, verbatim}
\usepackage{chronosys}
\usepackage{float}
\usepackage{subfig}
\usepackage{url}
\setlength{\textwidth}{7in} 
\setlength{\textheight}{9in}
\setlength{\oddsidemargin}{0in} 
\setlength{\evensidemargin}{0in}
\setlength{\topmargin}{-1.5cm}
\overfullrule=2cm
\usepackage[linktoc=all]{hyperref}
\usepackage{titlesec}
\hypersetup{colorlinks, citecolor=black, filecolor=black, linkcolor=black,urlcolor=black}
\usepackage[citestyle=verbose,bibstyle=numeric,sorting=ynt,backend=biber]{biblatex}
\usepackage[toc,page]{appendix}
\addbibresource{lit.bib}
\begin{document}
\setcounter{tocdepth}{2}
\linespread{1.3}
\tableofcontents

@@

A comparison of loan performance in different regimes allows me to consider two contrasting views of how an internet marketplace such as Prosper works.  One view, closely linked to Hayek, suggests that the marketplace would work best with minimal intervention from the market maker Prosper, so many individual lenders incorporate their private knowledge and information in the determination of the interest rate.   An opposing view would hold that individuals are not sophisticated or knowledgeable enough to assess loan risk, and some guidance or even interest rate setting from the market maker Prosper would improve loan performance.  The evidence on loan returns in the three regimes points quite clearly in favor of the market maker hypothesis, and against Hayek: as Prosper becomes more involved in the rate-setting process, the excess returns available using publicly available information approach 0, while aggregate investor returns exhibit higher mean and lower variance.

@@

Under all three regimes, lenders could choose to whom to lend and how much to
lend, allowing them to diversify across many loans, thereby decreasing the
variance of their returns. The minimum investment in a loan was \$25, with no
maximum, and lenders could hold as many loans as they wished.
Before a potential borrower's listing is posted on the website, Prosper authenticates their social security number, driver's
license and address, and pulls the borrower's credit history from Experian,
which includes credit score, total number of delinquencies, current
delinquencies, and many other historical credit variables.\footcite{freedman} 
In exchange, In exchange for verifying borrowers' identities and managing the
marketplace, Prosper charges a 0.5-3 percentage
point origination fee on each loan depending on the borrower's risk profile and
loan duration. 
This fee,usually 1\%, is added to the borrower's monthly payment.
Lenders are also charged a servicing fee of 0.5-1\%. There are no prepayment penalties.
This rate is lower than that charged by most traditional financial institutions, which have the higher costs of physical branches, and pay for human analysis of credit history.\footcite{hanson}
The P2PL model relies on lenders being able to infer the riskiness of borrowers: since Prosper gets fees from monthly payments, they too are incentivized to maximize investor returns, both for short term profit, and long term reputation.\newline

\iffalse
\begin{table}
  \renewcommand\arraystretch{1.4}\arrayrulecolor{LightSteelBlue3}
  \captionsetup{singlelinecheck=false, font=blue, labelfont=sc, labelsep=quad}
  \caption{Timeline}\vskip -1.5ex
  \begin{tabular}{@{\,}r <{\hskip 2pt} !{\foo}
    >{\raggedright\arraybackslash}p{5cm}}
    \toprule
    \addlinespace[1.5ex]
    1947 & AT and T Bell Labs develop the idea of cellular phones\\
    1968 & Xerox Palo Alto Research Centre envisage the 'Dynabook\\
    1971 & Busicom 'Handy-LE' Calculator\\
    1973 & First mobile handset invented by Martin Cooper\\
    1978 & Parker Bros. Merlin Computer Toy\\
    1981 & Osborne 1 Portable Computer\\
    1982 & Grid Compass 1100 Clamshell Laptop\\
    1983 & TRS-80 Model 100 Portable PC\\
    1984 & Psion Organiser Handheld Computer\\
    1991 & Psion Series 3 Minicomputer\\
  \end{tabular}
\end{table}
\fi
\iffalse
\startchronology
[startyear=2006,stopyear=2011,height=1ex,width=\hsize]
\chronoevent{0}{Jan 13 Quiz 1}
\chronoevent{8}{Jan 20 Quiz 2}
\chronoevent{36}{Feb 17 Quiz 6}
\chronoevent{36}{Intervention}
\stopchronology
\fi
In my sample, Prosper's origination volume, as shown in Figure \ref{fig:origTime}, peaked in the months before the shutdown, and then slowly resumed after the quiet period. Since 2011, when the sample ends, origination volume has eclipsed its 2008 peak.\newline
<<origTime, cache=FALSE, fig.cap='This chart shows two panels. The first panel shows the number of Loans funded on Prosper in each month of the sample. The second panel shows the total dollars lent to borrowers in each month of the sample.', fig.height=4>>=
g2 = group_by(nd, mgroup, regime) %>%
  summarise(Number_Of_Loans = length(mgroup), Dollars = sum(LoanOriginalAmount))
g2 = melt(g2,id.vars = c("mgroup",'regime'))

@@

<<catsTab, cache=FALSE>>=
catTab = filter(nd, purpose != "NA") %>% count(purpose, sort=TRUE) %>% mutate(Frequency = n/sum(n))
xtable(catTab, digits=3, label='catTab', caption = "Stated Reasons for Borrowing on Prosper (2006-2001). NAs removed")
@

@@

Second, whereas previously Prosper used only credit score to determine a borrower's Credit Grade (eg. AA), the site began offering a more advanced, multi variable formula to determine each borrower's credit grade for investors consideration. Finally, the site also partnered with FolioFN, a secondary market where lenders could liquidate their positions. Crucially, Prosper still had no direct role in setting interest rates, investors still used an auction, but with the extra information credit grades that more closely matched default probabilities. 
Besides credit score, Prosper began using (and still uses) the following ``Key variables" to determine its revamped credit grades: \footnote{CITE =  Estimated Loss Page}
\begin{itemize}
\item  TotalInquiries: The number of times potential lenders requested the borrowers' credit portfolio from the credit bureau, which is usually Experian or a competitor.
\item CurrentDelinquencies: The number of delinquent loans associated with the borrower.
\item Credit card utilization: how much of available credit the borrower uses on a monthly basis.
\item Number of open credit card accounts
\item Debt to income ratio
\item Loan payment performance on prior Prosper loans.
\end{itemize}
This Auction rate-setting mechanism continued for more than a year after the shutdown, until December 20, 2010.

Unsuprisingly the automated feature selection procedure in section 5, which
identifies the subset of variables that best predict default finds that
Credit Score, inquiries, delinquencies and number of credit cards are all useful
variables for predicting default.

\subsection{Prosper-rate Regime}
Under the current \emph{Prosper-rate} model, enacted December 21, 2010, lenders no longer determine the loan rate in an auction. Instead, Prosper uses the intelligent credit grades given to investors in the late auction regime to set interest rates on loans, with slightly less stringent under-writing standards to determine eligibility. The change was motivated by a desire to accelerate the listing process, so that investors could reinvest money more quickly. Moreover, towards the end of the late auction regime, auctions were arriving at interest rates that were almost completely determined by the credit score. Under the fixed-rate regime, investors choose whether or not to invest in a given loan, at the interest rate set by Prosper, so bids only consist of a dollar amount.\newline

\section{Literature}
\subsection{Efficient Markets Hypothesis}
In layman's terms, the efficient markets hypothesis (EMH), states that 'it is impossible to beat the market.' More specifically, Fama explains that market prices should always fully reflect publicly available information. Therefore, EMH rules out the possibility of strategies that earn excess returns using only publicly available information. \footcite{fama}
In the P2P lending context, this market price is the interest rate determined by
investors (or accepted by investors in the fixed-rate regime) at origination
time, and paid by the borrower, and returns are the cash flows received from borrowers divided by the amount invested.
(Fama, 1970) notes the dependence of most tests of market efficiency on a model of risk and return.
The models, including most famously, Sharpe's capital asset pricing model ('CAPM'), make two predictions:
First,  since an investor can always hold cash risklessly, the average return on risky investments must be positive. Second, securities with higher systematic risk earn higher average returns.\footcite{sharpe}

@@

Freedman and Jin (2008) are more focused on the failures of the auction regime,
Although the main focus of their paper is the role of soft information and
social networks on Prosper, they also study how investors learn to weight these
different variables, and confirm the finding of negative expected returns in the
early auction regime: ``lenders, especially those that joined
Prosper early, did not fully understand the market risk. We estimate, on
average, Lenders would have expected an internal rate of return of between
-.62\% and
1.38\% if they had correctly understood the risk distribution of Prosper
loans.''``Vigorous'' lender learning, they argue, explains the increase in interest
rates and borrower creditworthiness, and associated increase in lender
returns.\footcite{freedman}

@@

Rather differently than Hayek theorizes, the data suggest that when investors are left on their own
to set prices, markets are not nearly as efficient as the prices set by ``a
single mind'', that aggregates the data. Furthermore, as interest rate setting
becomes more centralized, interest rates are more sensitive to the hard credit variables
that actually predict default.
%t and other unused citations can be found in hypo.cut.tex

@@

\subsection{Variable Definitions for the Prosper Data}
<<sumTable,cache=FALSE>>=
nd$default.rate =  nd$defaulted
k4 = c('roi','default.rate', 'LenderYield','BorrowerRate','regime','LoanOriginalAmount','Investors','CreditScoreEst','DebtToIncomeRatio', 'InquiriesLast6Months')
ag4 <- aggregate(.~regime, nd[k4], function(x) c(mean = mean(x), sd = sd(x)))
d4 = t(do.call(data.frame, c(ag4, check.names = FALSE)))
d5 = data.frame(d4)
names(d5) = d4[1,]
d6 = d5[-1,]
xtable(d6,caption='Means and Standard Deviations for Relevant Variables by Regime.',digits=3, label='sumStats')
@
For each loan, Prosper offers 63 variables, I now define an important subset.
\begin{itemize}
\item BorrowerRate: Annual interest rate paid by the borrower
\item TotalCreditLinespast7years: Number of credit lines, including cards and loans, used by the borrower in the seven years before the prosper loan was originated.
\item Inquirieslast6Months: Number of times a lender inquired about the Borrower's credit report, in the 6 months before the loan was originated. Traditional lenders make inquiries when a borrower applies for a loan, so this serves as a proxy for the number of times someone has borrowed in the past 6 months.
\item CurrentDelinquencies: Number of loans on which the borrower is delinquent at the origination date.
\item DelinquenciesLast7Years: The number of loans which the borrower has not repaid fully over the 7 years before the origination date.
\item MonthlyLoanPayment: The borrower's scheduled monthly payment on the Prosper loan.
\item CreditScoreRangeLower: The lower bound of the credit score range (which
  bounds his raw credit score) , that bounds a borrower's credit grade, as
  provided by Experian. Prosper provides the range instead of the raw score to
  preserve borrower anonymity.
\item CreditScoreEst: The average of the lower and upper bounds of the credit score range. 
\item PublicRecordsLast10Years: Number of public records of borrowing activity in the past 10 years at the origination date. 
\item DebtToIncomeRatio: The debt to income ratio of the borrower at the origination date. This value is capped at 10.01 (any debt to income ratio larger than 1000\% will be coded as 1001\%).
\item TotalProsperPaymentsBilled: Number of on time payments the borrower made on Prosper loans at the time they created the listing. This value is 0 if the borrower had no prior loans.
\item LenderYield: The lender's Yield on the loan. Equal to the Interest rate minus servicing fee.
\item BorrowerRate: The borrower's interest rate for the loan.
\item Investors: The number of investors that funded the loan.
\item *.sc: The Variable * was scaled to have 0 mean and unit variance. (found only in regression results)
\end{itemize}

@@

I do not discount or annualize ROIs, because in the period of low interest
rates, the difference between default and repayment is large enough that a few
percentage points don't matter: loans that default average -46\% ROI, loans that
repay average + 24\% ROI. Furthermore, my ROI results are very similar as Freedman
and Jin's, who use discounting.\footcite{freedman} %TODO(SS): CITE FED RATES

@@

I find the following: (1) the mean ROI of early auction regime loans is
negative, (2) in the early auction regime, average ROI is lower for higher risk
loans, (3) interest rates are insufficiently sensitive to risk in the early
auction period and (4) investors made sub-optimal choices, given the information
they had at their disposal, and (5) In the period when Prosper intervened,
interest rates and investor returns were much higher, and part of the increase
is likely due to Prosper's intervention.

These findings suggests that although investors were unlucky in
the early auction regime, they made avoidably bad decisions that hurt their
returns. More specifically,
Finding (1) is inconsistent with the EMH prediction that risky assets like Prosper loans should offer investors a positive expected return.\footnote{Otherwise, investors would just hold cash.}
Finding (2) is inconsistent with the prediction that riskier loans should have higher expected returns to compensate investors for taking risks.
Finding (3) describes the sub-optimal investor decisions that led to (1) and (2), in response to arguments that investors chose an unlucky time period.
Finding (4) is inconsistent with Hayek's notion that a free market will arrive
at accurate prices that reflect the information available to market
participants,without the meddling of a central planner.

@@

ggplot(rateDiff, aes(x=CreditScoreEst, y=meanRoi,size=num,color=regime))+
  geom_line() +
  theme_bw() +
  xlim(500,900) + 
  scale_y_continuous(labels=percent, limits=c(-.3,NA)) +
  labs(title='Estimated Credit Score vs. ROI',x= 'Estimated Credit Score'
       , y='Mean ROI',color='Regime',size='Number of Loans')
#Total Inquiries, other cluster vars.
@

Figure \ref{fig:roivscore} shows that lower credit score loans earn lower returns in the early auction regime, inconsistent with the EMH prediction of a risk, reward tradeoff.
The expected pattern is shown in later two regimes' curves: riskier loans earn higher average returns. Figure \ref{fig:roivscore} also reveals early auction regime returns are worse than the other two regimes' returns for each bucket, but that the gap is much greater for riskier loans. Again, time period may be a confounding factor, as riskier investments often perform worse in bad economic times, a possibility I discuss later.
We now examine default rates and interest rates, the two features that explain the variation inreturns.

\subsection{Interest Rates and Defaults Across regimes}
<<ratevtime,cache=FALSE,fig.height=3, fig.cap="The blurred lines show the median interest rate of loans originated each month. The solid lines represent the median rate for loans to borrowers in the 640-660 credit score bucket for each month. Loans funded after the shutdown are associated with higher rates after controlling for borrower credit score bucket.">>=
ggplot() +
  geom_line(aes(x=mmeans$mgroup, y=mmeans$mRate, size=mmeans$numLoans, color=mmeans$regime),alpha=.5) + 
  geom_line(aes(x=mm3$mgroup, y=mm3$mRate, size=mm3$numLoans, color=mm3$regime)) + 
  theme_bw() +
  labs(title='Interest Rates Over Time (read caption)', color='Regime',size='Number of Loans',x='Origination Month',y='Mean Rate')+
  scale_y_continuous(label=percent, limits=c(.1, .33))
@

@@

<<defscore.tab,fig.height=3,cache=FALSE,fig.pos='H'>>=
a1 = filter(d2, regime=='Auction-Early')[c('CreditScoreEst','dr')]
colnames(a1) = c('CreditScoreEst','Auction-Early')
a2 = filter(d2, regime=='Auction-Late')[c('CreditScoreEst','dr')]
colnames(a2) = c('CreditScoreEst','Auction-Late')
a3 = filter(d2, regime=='Prosper-rate')[c('CreditScoreEst','dr')]
colnames(a3) = c('CreditScoreEst','Prosper-rate')
ds.tab = left_join(a1, inner_join(a2,a3, by='CreditScoreEst'), by='CreditScoreEst') %>% filter(CreditScoreEst > 600 & CreditScoreEst < 880)
colnames(ds.tab)[1] = 'Credit Score Bucket'
xtable(ds.tab, label='ds.tab', caption='Default rates for borrowers in each credit score for each regime. Credit score buckets are denoted by their midpoint. EG, the 600-619 bucket is in the first row, 609.5. Each column then represents the default rates borrowers in the (column name) regime, in the (row-name) credit bucket. Default rates are much higher for the early auction regime, especially among risky borrowers. This data underlies the previous figure. Early auction regime borrowers in buckets below 500 have even higher default rates.')
@
Lower sensitivity to credit scores, and total disregard for differences in
credit score below 600 did not work out well for early auction regime investors,
the relationship between credit score and default probability was even stronger
in that regime than in the other regimes. 
Figure \ref{fig:defvscore} presents the relationship between credit score bucket
and default rate, and shows that the probability of default decreases
(increases) with higher (lower) credit score, for all three regimes.
In the early auction regime, a 100 point increase (5 buckets) in credit score is
associated with a 15\% reduction in default probability. In the late auction
regime, a 100 point increase is associated with a 7\% reduction in default
probability. In the Prosper-rate regime, an 100 point increase is associated
with a 12\% increase in default probability. The data that underly Figure \ref{fig:defvscore}
can be viewed in tabular form in Table \ref{ds.tab}.

@@

Since we do not know Propser's exact grading formula, it may include information
that was not previously available to borrowers. Table \ref{ratefit2} addresses this argument, by running the same procedure as Table \ref{ratefit}, but with only credit score bucket as an independent variable. The direction of the results is the same: Prosper's change to more informative grading did not only lead to a larger role for those grades in investors' interest rates, but increased the weight of credit score bucket, a variable that investors had access to in the same format in both periods. If the crowd of investors were as wise as Hayek theorizes, Prosper's repackaging of multiple variables into a more informative credit grade should not affect the weight of hard information in investors' risk estimates. Furthermore, Figure \ref{fig:defvscore} makes is clear that investors would have experienced fewer defaults, and thus earned higher returns, if they had used credit buckets more heavily in their risk assessments, given the poor returns associated with low credit borrowers. We can thus interpret some of the poor returns of the early auction regime to a lower level of reliance on hard information.

In the next sub section, I estimate the impact of Prosper's other interventions, which ban low-credit borrowers.
\subsection{Impact of Subprime Borrower Ban}
<<csvtime,cache=FALSE,fig.height=3, fig.cap="This figure presents average estimated credit score of loans originated each month. The average credit score (using bucket midpoint) increases from 653 in the early auction regime to 714 in the late auction regime, before returning to 705 in the Prosper-rate regime. Grey clouds indicate one standard deviation.">>=
ggplot(mmeans, aes(x=mgroup, y=mcs, color=regime)) +
  geom_point(aes(size=numLoans))+
  geom_smooth(aes(ymax=mcs.up, ymin=mcs.down),stat='identity') +
  #geom_line() +
  theme_bw() +
  labs(title='Borrower Credit Scores Over Time', size='Number of Loans',x='Origination Month', y='Mean Borrower Credit Score')
##TODO(ADD ERROR BARS)
@
We now turn to Prosper's policy changes, which restrict the types of borrowers who investors are allowed to lend to. An auction perspective predicts that the crowd of lenders should be able to effectively screen all candidates, not just low risk candidates. On multiple occasions, however, Prosper has intervened to restrict the borrower pool to high credit borrowers, and thereby improved investor returns. Figure \ref{fig:csvtime} shows the mean borrower credit score for each month in the sample, with the grey clouds showing one standard deviation bounds. Overall, for the three samples in full, the average borrower credit scores increases from 653.2 before the shutdown to 714.2 after the shutdown, then dips to 705 for the Prosper-rate period.
For the first 10 months of Prosper's history, investors choose to lend to borrowers with an average estimated credit score of 609. Prosper's policy change, on March 1, 2007, that prevented loans to ``low-credit borrowers," defined as those with credit score below 520, showed an immediate positive impact on investor decision making and returns.
